Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2021-05-28 ms/Thesis/thesis.tex
invasion with general vertical transmission

%%% Figure: Invasion
\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{../PRSB_figures/fig_s1.pdf}
  \caption{
  \textbf{Reduction principle for interaction-transmission association.} 
  Consecutive fixation of modifier alleles that reduce interaction-transmission association $\alpha$ in numerical simulations of evolution with two modifier alleles (Eq.~D1).
  When an invading modifier allele is established in the population (frequency > 99.95\%), a new modifier allele that reduces interaction-transmission association by 5\% is introduced (at initial frequency 0.5\%).
  \textbf{(a)}~The frequency of the cooperative phenotype $A$ over time.
  \textbf{(b)}~The frequency of the invading modifier allele $m$ over time. 
  \textbf{(c)}~The population mean fitness ($\bar{w}$) over time.
  Here, $c = 0.05$, $b=1.3$, $T_A=0.4<T_B=0.7$, initial interaction-transmission association $\alpha_1=0.7$, lower interaction-transmission association threshold $a_2=0.605$.  
  }
  \label{fig:invasion}
\end{figure}
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Population structure}
All the simulations in this section were made by Ohad Lewin-Epstein from Tel Aviv University.
Interaction-transmission association may also emerge from population structure.
Consider a  population colonizing a two-dimensional grid of size 100-by-100, where each site is inhabited by one individual, similarly to the model of \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype: either cooperator, $A$, or defector, $B$.
Initially, each site in the grid is randomly colonized by either a cooperator or a defector, with equal probability.
In each generation, half of the individuals are randomly chosen to "initiate" interactions, and these
initiators interact with a random neighbor (i.e. individual in a neighboring site) in a prisoners' dilemma game (\autoref{fig:illustration}c) and a random neighbor (with replacement) for horizontal cultural transmission (\autoref{fig:illustration}b).
The expected number of each of these interactions per individual per generation is one, but the realized number of interactions can be zero, one, or even more than one, and in every interaction both individuals are affected, not just the initiator.
The effective interaction-transmission association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/M$, where $M$ is the number of neighbors.
On an infinite grid, $M=8$ (i.e. Moore neighbourhood~\citep{moore1962machine}), but on a finite grid $M$ can be lower in neighbourhoods close to the grid border.
As before, $T_A$ and $T_B$ are the probabilities of successful horizontal transmission of phenotypes $A$ and $B$, respectively.

>>>>

We can do similiar analysis for the general case where $0<v<1$ as we did in Appendix~\ref{sec:appendixD}.
First we found the value of the characteristic polynomial at 1, $R(1)$ using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics.
\begin{equation} \label{eq:value_at_1}
  \begin{aligned}
  R(1) &= \frac{c v \hat{p}^*[T_Ab\hat{p}^{*2}(v\alpha_2-1)-2T_Ab\hat{p}^*v\alpha_2 + T_Ab\hat{p}^*(1+v\alpha_2)]}{b\hat{p}(b\hat{p}^*-2\hat{p}^*+2)+c\hat{p}^*(c\hat{p}^*-2)+1} \\
 & + \frac{c v \hat{p}^*[T_Ac\hat{p}^{*2}(1-v)+T_Ac\hat{p}^*(v-1)-T_A\hat{p}^*(1-c)+T_A]}{b\hat{p}(b\hat{p}^*-2\hat{p}^*+2)+c\hat{p}^*(c\hat{p}^*-2)+1} \\
  &+ \frac{c v \hat{p}^*[T_Bb\hat{p}^{*2}(1-v\alpha_2)+T_Bb\hat{p}^2(v\alpha_2-1)+T_Bc\hat{p}^{*2}(v-1)]}{b\hat{p}(b\hat{p}^*-2\hat{p}^*+2)+c\hat{p}^*(c\hat{p}^*-2)+1}\\
  &+ \frac{c v \hat{p}^*[T_Bc\hat{p}^*(1-v) + T_Bcv(1-\hat{p}^*+T_B(\hat{p}^*-1)+cv(\hat{p}^*-1))]}{b\hat{p}(b\hat{p}^*-2\hat{p}^*+2)+c\hat{p}^*(c\hat{p}^*-2)+1}
  \end{aligned}
\end{equation}
Similarly to what we did in Appendix~\ref{sec:appendixD}, we should find when $R(1)<0$. However, when adding the vertical transmission into the system the math is more complicated. 
Therefore, we will use a different technique. We know that when $\alpha_1 = \alpha_2$, $R(1)=0$. 
% DC: Yoav should I proof the statement above?
Now, let's assume that $\alpha_2 = \alpha_1 + \epsilon$. The derivative of $R(1)$ by $\epsilon$ will give us better understand about the sign of $R(1)$ when $\alpha_2<\alpha_1$.
\begin{equation} \label{eq:value_at_1_derivative}
  \frac{\partial R(1)}{\partial \epsilon} = \frac{c b  v^2 \hat{p}^*[T_A\hat{p}^{*2}-2T_A\hat{p}^* + T_A-T_B\hat{p}^{*2}+T_B\hat{p}^*]}{b\hat{p}(b\hat{p}^*-2\hat{p}^*+2)+c\hat{p}^*(c\hat{p}^*-2)+1} 
\end{equation}
We can simplify \autoref{eq:value_at_1_derivative}
\begin{equation} \label{eq:value_at_1_derivative_simplify}
  \frac{\partial R(1)}{\partial \epsilon} = \frac{c b v^2 \hat{p}^*[(T_A-T_B)\hat{p}^{*2}+(T_B-2T_A)\hat{p}^* + T_A]}{(b\hat{p}^*-c\hat{p}^*)^2+2\hat{p}^*(b-c)+1} 
\end{equation}
Since $b>c$ the denominator is always positive. The numerator is quadratic polynomial of $\hat{p}^*$ with the following roots:
\begin{equation} \label{eq:numerator_roots}
  \hat{p}^*_1 = \frac{T_A}{T_A-T_B} \hat{p}^*_2 = 1
\end{equation}
$\hat{p}^*_1 < 0$ since $T_B>T_A$ and since the quadratic polynomial has negative leading coefficient then the numerator is positive for every $\hat{p}^*_1<\hat{p}^*<1$ which is always true.
We found that derivative of $R(1)$ by $\epsilon$ is positive for every $\epsilon$. 
Therefore, $R(1)$ grows as $\epsilon$ grows, or in other words as $\alpha_2-\alpha_1$ grows.
Thus, $R(1)<0$ if and only if $\alpha_2<\alpha_1$. Therefore, different value of vertical transmission does not change the evolution of social-interaction association and invasion is possible when $\alpha_2<\alpha_1$.
% DC: can I say if and only if

%%% Figure: Invasion
\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{../PRSB_figures/fig_s1.pdf}
  \caption{
  \textbf{Reduction principle for interaction-transmission association.} 
  Consecutive fixation of modifier alleles that reduce interaction-transmission association $\alpha$ in numerical simulations of evolution with two modifier alleles (Eq.~D1).
  When an invading modifier allele is established in the population (frequency > 99.95\%), a new modifier allele that reduces interaction-transmission association by 5\% is introduced (at initial frequency 0.5\%).
  \textbf{(a)}~The frequency of the cooperative phenotype $A$ over time.
  \textbf{(b)}~The frequency of the invading modifier allele $m$ over time. 
  \textbf{(c)}~The population mean fitness ($\bar{w}$) over time.
  Here, $v=1$, $c = 0.05$, $b=1.3$, $T_A=0.4<T_B=0.7$, initial interaction-transmission association $\alpha_1=0.7$, lower interaction-transmission association threshold $a_2=0.605$.  
  }
  \label{fig:invasion}
\end{figure}
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Population structure}
All the simulations in this section were made by Ohad Lewin-Epstein from Tel Aviv University.
Interaction-transmission association may also emerge from population structure.
Consider a  population colonizing a two-dimensional grid of size 100-by-100, where each site is inhabited by one individual, similarly to the model of \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype: either cooperator, $A$, or defector, $B$.
Initially, each site in the grid is randomly colonized by either a cooperator or a defector, with equal probability.
In each generation, half of the individuals are randomly chosen to "initiate" interactions, and these
initiators interact with a random neighbor (i.e. individual in a neighboring site) in a prisoners' dilemma game (\autoref{fig:illustration}c) and a random neighbor (with replacement) for horizontal cultural transmission (\autoref{fig:illustration}b).
The expected number of each of these interactions per individual per generation is one, but the realized number of interactions can be zero, one, or even more than one, and in every interaction both individuals are affected, not just the initiator.
The effective interaction-transmission association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/M$, where $M$ is the number of neighbors.
On an infinite grid, $M=8$ (i.e. Moore neighbourhood~\citep{moore1962machine}), but on a finite grid $M$ can be lower in neighbourhoods close to the grid border.
As before, $T_A$ and $T_B$ are the probabilities of successful horizontal transmission of phenotypes $A$ and $B$, respectively.
